---
title: "Hurdle GAM for deconvolved calcium events"
subtitle: "First task of the first mouse"
author: "LAPD"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

params <- list(
  holdout_seconds = 30,
  threshold_method = "three_sd",
  n_neurons = NULL,
  nthreads = 2,
  max_diagnostic_rows = 500000,
  concurvity_rows = 20000,
  cache_trial = TRUE
)

knitr::opts_chunk$set(fig.align = "center")

required_packages <- c("data.table", "ggplot2", "knitr", "mgcv")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the following packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(ggplot2)

# Quarto normally evaluates relative to this file's directory. The fallback
# also permits executing the chunks interactively from the repository root.
root_candidates <- c("..", ".")
root_matches <- root_candidates[
  file.exists(file.path(root_candidates, "all_trials.RDS"))
]
if (!length(root_matches)) stop("Could not locate all_trials.RDS.")
project_root <- normalizePath(root_matches[1L])

source(file.path(project_root, "script", "data_preparation.R"))
source(file.path(project_root, "script", "hurdle_gam.R"))
source(file.path(project_root, "script", "model_diagnostics.R"))

theme_set(theme_minimal(base_size = 12))

get_metric <- function(results, component, predictor, metric) {
  value <- results$value[
    results$component == component &
      results$predictor == predictor &
      results$metric == metric
  ]
  if (length(value) != 1L) stop("Could not identify a unique metric value.")
  value
}
```

# Objective and data

The response is the deconvolved spike estimate $S_{it}$ for neuron $i$ at calcium frame $t$. It is a sparse, nonnegative estimate produced from calcium fluorescence rather than a directly observed action-potential train. We therefore call a threshold crossing a **calcium event**, not a known physiological spike.

The first mouse is `M3411`, and its first task is `trial1`, a triangular-arena trial. It contains 462 neurons and approximately 10,870 calcium frames at 15 Hz. Behavioral positions were recorded at approximately 30 Hz. The timestamps stored under `neuron$time` describe the original acquisition clock and are not one-to-one with the columns of `S`; consequently, we construct a calcium-frame time grid over the recording interval and linearly interpolate behavior onto it.

The source RDS is a large nested object. On the first execution, the following chunk reads it and saves only the selected trial in `analysis/cache`. Later renders use that much smaller cache. Delete the cached file if the source data change.

```{r}
#| label: load-trial

cache_path <- if (isTRUE(params$cache_trial)) {
  file.path(project_root, "analysis", "cache", "M3411_trial1.rds")
} else {
  NULL
}

trial_data <- load_trial_data(
  rds_path = file.path(project_root, "all_trials.RDS"),
  mouse = 1L,
  trial = 1L,
  cache_path = cache_path
)

data.frame(
  mouse = trial_data$mouse_name,
  trial = trial_data$trial_name,
  arena = trial_data$shape,
  neurons = nrow(trial_data$neuron$S),
  calcium_frames = ncol(trial_data$neuron$S),
  behaviour_observations = nrow(trial_data$behav$position)
) |>
  knitr::kable(caption = "Selected recording")
```

## Temporal split and event definition

The last `r params$holdout_seconds` seconds are held out. This is deliberately harder and more realistic than randomly holding out individual rows: nearby frames are strongly dependent, and a random split would leak almost identical behavioral states into both sets.

For the default event definition, the threshold for neuron $i$ is estimated from training frames only:

$$
c_i = 3\,\operatorname{SD}_{t\in\mathcal T_{\mathrm{train}}}(S_{it}),
\qquad
Y_{it}=\mathbb{1}(S_{it}>c_i).
$$

This follows the three-standard-deviation rule described in the experimental paper. Both the threshold and all covariate standardization constants are calculated without using the terminal holdout. The conditional amplitude response is the original $S_{it}$ on frames for which $Y_{it}=1$.

```{r}
#| label: prepare-data

all_neurons <- seq_len(nrow(trial_data$neuron$S))
neuron_indices <- if (is.null(params$n_neurons)) {
  all_neurons
} else {
  head(all_neurons, as.integer(params$n_neurons))
}

analysis_data <- prepare_hurdle_analysis_data(
  trial_data = trial_data,
  holdout_seconds = params$holdout_seconds,
  threshold_method = params$threshold_method,
  neuron_indices = neuron_indices
)

train_data <- analysis_data$train
test_data <- analysis_data$test

split_summary <- rbind(
  data.frame(
    set = "Training",
    frames = data.table::uniqueN(train_data$frame),
    observations = nrow(train_data),
    events = sum(train_data$event),
    event_rate = mean(train_data$event),
    start_sec = min(train_data$time_sec),
    end_sec = max(train_data$time_sec)
  ),
  data.frame(
    set = "Terminal holdout",
    frames = data.table::uniqueN(test_data$frame),
    observations = nrow(test_data),
    events = sum(test_data$event),
    event_rate = mean(test_data$event),
    start_sec = min(test_data$time_sec),
    end_sec = max(test_data$time_sec)
  )
)
knitr::kable(split_summary, digits = 4, caption = "Analysis split and event prevalence")
```

## Behavioral covariates and arena boundary

Let $(x_t,y_t)$ be the interpolated mouse position, $d_t$ its Euclidean distance from the nearest estimated triangle edge, and $v_t$ its smoothed displacement per second in the coordinate system of the tracking camera. The triangle is estimated from training positions only. Values are standardized using training means and standard deviations before fitting.

Distance to the boundary is a deterministic spatial summary and is therefore related to the bivariate position term below. We retain both to estimate a directly interpretable boundary effect after accounting for broader spatial variation, but inspect concurvity after fitting. A high concurvity value means the separate effects should not be interpreted in isolation.

```{r}
#| label: arena-check
#| fig-cap: "Aligned trajectory and triangle boundary estimated from training positions. The red segment is the terminal holdout."

boundary_df <- boundary_dataframe(analysis_data$boundary)
frames <- analysis_data$frames

ggplot(frames, aes(x, y)) +
  geom_path(aes(color = set), linewidth = 0.25, alpha = 0.65) +
  geom_path(data = boundary_df, inherit.aes = FALSE, aes(x, y),
            linewidth = 1, color = "black") +
  scale_color_manual(values = c(train = "grey55", test = "#D55E00"),
                     labels = c(train = "Training", test = "Terminal holdout")) +
  coord_equal() +
  labs(x = "Position x", y = "Position y", color = NULL)
```

# Statistical model

The hurdle model has two components. First, event occurrence follows a binomial GAM:

$$
Y_{it}\mid p_{it}\sim\operatorname{Bernoulli}(p_{it}),
$$

$$
\operatorname{logit}(p_{it})=
\beta_0+b_i+f_{xy}(x_t,y_t)+f_d(d_t)+f_v(v_t)+f_T(T_t).
$$

Here $b_i\sim N(0,\sigma_b^2)$ is a penalized neuron random intercept. The bivariate smooth $f_{xy}$ captures population-level spatial variation that is not reducible to distance from the wall. The univariate smooths describe boundary distance, movement speed, and gradual change over trial time. The time term can capture drift or behavioral nonstationarity, but predictions in the terminal holdout necessarily extrapolate this term slightly beyond the training range.

Second, the positive deconvolution amplitude follows a Gamma GAM with a log link:

$$
S_{it}\mid(S_{it}>c_i),\mu_{it},\phi
\sim\operatorname{Gamma}(\mu_{it},\phi),
$$

$$
\log(\mu_{it})=
\alpha_0+u_i+g_{xy}(x_t,y_t)+g_d(d_t)+g_v(v_t)+g_T(T_t),
$$

where the Gamma distribution is parameterized by its mean and dispersion. The two linear predictors have the same structure but are estimated independently. They need not have the same smooth shapes or smoothing strengths.

The implied conditional mean of the thresholded response

$$
S^*_{it}=S_{it}\mathbb{1}(S_{it}>c_i)
$$

is

$$
\operatorname{E}(S^*_{it}\mid\mathbf x_t,i)=p_{it}\mu_{it}.
$$

This is a marginal regression model. It does not assert independence of the underlying neurons and does not identify directed neuronal interactions. The blocked holdout partly protects predictive assessment from temporal dependence, but conventional GAM standard errors may still be optimistic.

# Model fitting

The models are fitted with `mgcv::bam()`, using fast restricted maximum likelihood, discretized covariates, shrinkage selection, and low-rank smooth bases. These choices are intended to make the pooled model practical on a laptop. The positive component uses only detected training events and is therefore much smaller than the occurrence component.

```{r}
#| label: fit-model

formulas <- default_hurdle_formulas(
  position_k = 50L,
  border_k = 8L,
  speed_k = 8L,
  time_k = 12L
)

hurdle_fit <- fit_hurdle_gam(
  train_data,
  formulas = formulas,
  nthreads = params$nthreads
)

hurdle_fit
```

```{r}
#| label: model-summary

model_summary <- hurdle_model_summary(hurdle_fit)
model_summary |>
  transform(
    adjusted_r_squared = round(adjusted_r_squared, 4),
    deviance_explained = round(deviance_explained, 4)
  ) |>
  knitr::kable(caption = "Model-level in-sample summaries")
```

The two components behave very differently. The occurrence model explains only
`r sprintf("%.2f%%", 100 * model_summary$deviance_explained[model_summary$component == "Event occurrence"])`
of the binomial deviance, with adjusted $R^2$ equal to
`r sprintf("%.4f", model_summary$adjusted_r_squared[model_summary$component == "Event occurrence"])`.
Some behavioral modulation is present, but most frame-to-frame event variation
is not explained by the shared covariates. In contrast, the positive-amplitude
model explains
`r sprintf("%.1f%%", 100 * model_summary$deviance_explained[model_summary$component == "Positive amplitude"])`
of its in-sample deviance. This apparently strong amplitude fit must still be
checked out of sample: it can reflect stable differences among neurons captured
by the random intercepts rather than stable behavioral effects.

Deviance explained is an in-sample descriptive measure, not an estimate of
generalization. The very low event prevalence also makes adjusted $R^2$ less
informative than log loss, calibration, and blocked validation.

```{r}
#| label: concurvity

occurrence_concurvity <- sampled_concurvity(
  hurdle_fit$occurrence,
  newdata = train_data,
  max_rows = params$concurvity_rows,
  exclude_pattern = "neuron"
)
concurvity_table <- data.frame(
  term = colnames(occurrence_concurvity),
  worst = as.numeric(occurrence_concurvity["worst", ]),
  observed = as.numeric(occurrence_concurvity["observed", ]),
  estimate = as.numeric(occurrence_concurvity["estimate", ])
)
knitr::kable(
  concurvity_table,
  digits = 3,
  caption = "Concurvity diagnostics for the occurrence model"
)
```

The diagnostic uses a fixed uniform sample of
`r format(attr(occurrence_concurvity, "sample_rows"), big.mark = ",")`
training rows. Computing `mgcv::concurvity()` directly would construct the
complete multi-million-row model matrix and can overflow its internal integer
dimension calculation. The neuron random-effect term is omitted here because
the diagnostic of interest is overlap among the behavioral smooths. Every
neuron is observed at every frame, so neuron identity is balanced with respect
to these frame-level covariates.

The fitted result confirms a serious specification problem. The observed
concurvity for `s(border_z)` is
`r sprintf("%.3f", concurvity_table$observed[concurvity_table$term == "s(border_z)"])`,
meaning that its fitted contribution is almost completely reproducible from
the remaining terms, principally the two-dimensional position smooth. Its
separate shape should therefore not be interpreted as an independent boundary
effect. Speed has moderate observed concurvity
(`r sprintf("%.3f", concurvity_table$observed[concurvity_table$term == "s(speed_z)"])`),
while the fitted bivariate-position contribution is less redundant on the
observed measure
(`r sprintf("%.3f", concurvity_table$observed[concurvity_table$term == "s(x_z,y_z)"])`).
A more defensible comparison would fit position-only and boundary-only spatial
models and compare blocked predictive performance, rather than placing both in
one linear predictor.

# In-sample goodness of fit

To keep prediction and plotting memory bounded, proper scores and calibration are calculated on a fixed uniform sample of at most `r format(params$max_diagnostic_rows, big.mark = ",")` training rows. Uniform sampling preserves the event prevalence. The fitted-model deviance summaries above use the complete training set.

The comparison baseline assigns each neuron its smoothed training event rate and its mean positive training amplitude; it contains no position, speed, boundary, or time information. A useful GAM should outperform this neuron-only baseline.

```{r}
#| label: training-predictions

baseline_fit <- fit_neuron_baseline(train_data)
train_diagnostic <- sample_diagnostic_rows(
  train_data,
  max_rows = params$max_diagnostic_rows
)
train_diagnostic <- add_hurdle_predictions(hurdle_fit, train_diagnostic)
train_diagnostic <- add_baseline_predictions(train_diagnostic, baseline_fit)

training_metrics <- evaluate_hurdle_predictions(
  train_diagnostic,
  set_label = "Training sample"
)
knitr::kable(
  training_metrics,
  digits = 5,
  caption = "In-sample predictive fit on a uniform diagnostic sample"
)
```

In sample, the behavioral GAM raises event AUC from
`r sprintf("%.3f", get_metric(training_metrics, "Event occurrence", "Neuron baseline", "AUC"))`
for the neuron-only baseline to
`r sprintf("%.3f", get_metric(training_metrics, "Event occurrence", "Hurdle GAM", "AUC"))`.
This is real but modest ranking improvement. Log loss falls only from
`r sprintf("%.5f", get_metric(training_metrics, "Event occurrence", "Neuron baseline", "log loss"))`
to
`r sprintf("%.5f", get_metric(training_metrics, "Event occurrence", "Hurdle GAM", "log loss"))`,
and the Brier scores are indistinguishable at the displayed precision because
non-events dominate this response. The positive-amplitude RMSE improves from
`r sprintf("%.2f", get_metric(training_metrics, "Positive amplitude", "Neuron baseline", "RMSE"))`
to
`r sprintf("%.2f", get_metric(training_metrics, "Positive amplitude", "Hurdle GAM", "RMSE"))`.
The combined thresholded-amplitude RMSE changes by less than
`r sprintf("%.2f%%", 100 * (get_metric(training_metrics, "Thresholded amplitude", "Neuron baseline", "RMSE") - get_metric(training_metrics, "Thresholded amplitude", "Hurdle GAM", "RMSE")) / get_metric(training_metrics, "Thresholded amplitude", "Neuron baseline", "RMSE"))`.
Thus, even before looking at the holdout, the extra behavioral terms add only a
small amount beyond persistent neuron-to-neuron differences.

## Event calibration

A calibrated event model has mean observed event frequency close to mean predicted probability. The bins below contain approximately equal numbers of neuron-frame observations. Because events are rare, both axes focus on small probabilities.

```{r}
#| label: training-calibration
#| fig-cap: "In-sample calibration of calcium-event probabilities."

train_calibration <- calibration_table(
  train_diagnostic$event,
  train_diagnostic$event_probability,
  n_bins = 10L
)

ggplot(train_calibration, aes(predicted, observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey45") +
  geom_point(aes(size = n), color = "#0072B2") +
  geom_line(color = "#0072B2") +
  coord_equal() +
  labs(x = "Mean predicted probability", y = "Observed event frequency", size = "Rows")
```

The in-sample calibration points lie close to the identity line. This indicates
reasonable average calibration over the training period, but it does not imply
that the model captures event timing or transfers to later frames.

## Population activity over an in-sample interval

The next plot uses the last minute of the training period and averages across neurons within one-second bins. It checks whether the fitted marginal model follows broad changes in population event rate without requiring it to reproduce individual event times exactly.

```{r}
#| label: training-time-fit
#| fig-cap: "Observed and fitted population event rates during the last minute of training."

training_end <- max(train_data$time_sec)
train_tail <- train_data[train_data$time_sec >= training_end - 60, ]
train_tail <- add_hurdle_predictions(hurdle_fit, train_tail)
train_time <- population_time_summary(train_tail, bin_seconds = 1)
train_time_long <- rbind(
  data.frame(time_sec = train_time$time_sec, rate = train_time$observed_event_rate,
             series = "Observed"),
  data.frame(time_sec = train_time$time_sec, rate = train_time$predicted_event_rate,
             series = "Fitted")
)

ggplot(train_time_long, aes(time_sec, rate, color = series)) +
  geom_line(linewidth = 0.7) +
  scale_color_manual(values = c(Observed = "black", Fitted = "#0072B2")) +
  labs(x = "Trial time (seconds)", y = "Mean event rate", color = NULL)
```

The fitted curve follows some slow changes in population activity but is much
smoother than the observed series. It misses sharp peaks and troughs, suggesting
unmodeled short-timescale population states, coactivation, or serial dependence.

## Spatial goodness of fit

Observed and fitted event rates are averaged within spatial bins. Empty or rarely visited bins are omitted from visual emphasis by mapping their sample size to opacity.

```{r}
#| label: spatial-fit
#| fig-cap: "Observed and fitted in-sample event rates across the arena."

spatial_fit <- spatial_fit_summary(train_diagnostic, n_bins = 15L)
spatial_fit_long <- rbind(
  data.frame(spatial_fit[, c("x", "y", "n")], rate = spatial_fit$observed,
             quantity = "Observed"),
  data.frame(spatial_fit[, c("x", "y", "n")], rate = spatial_fit$predicted,
             quantity = "Fitted")
)

ggplot(spatial_fit_long, aes(x, y, color = rate, alpha = n)) +
  geom_point(size = 3) +
  facet_wrap(~quantity) +
  scale_color_viridis_c() +
  scale_alpha(range = c(0.25, 1), guide = "none") +
  coord_equal() +
  labs(x = "Position x", y = "Position y", color = "Event rate")
```

The fitted map reproduces the broad population-average spatial gradient but
smooths away much of the local observed variation. More importantly, averaging
over 462 neurons can make this plot look satisfactory even when individual
place fields are poorly modeled: different neurons can prefer different
locations while the current model forces them to share one spatial surface.

# Out-of-sample terminal prediction

No part of the terminal holdout was used to estimate event thresholds, the arena boundary, scaling constants, smooths, random effects, or baseline rates. The test therefore asks whether relationships learned earlier in the same task continue to predict the final `r params$holdout_seconds` seconds. It is a within-neuron, within-task assessment; it does not establish generalization to a new neuron, arena, trial, or mouse.

```{r}
#| label: test-predictions

test_predictions <- add_hurdle_predictions(hurdle_fit, test_data)
test_predictions <- add_baseline_predictions(test_predictions, baseline_fit)

test_metrics <- evaluate_hurdle_predictions(
  test_predictions,
  set_label = "Terminal holdout"
)
knitr::kable(
  test_metrics,
  digits = 5,
  caption = "Out-of-sample performance on the terminal holdout"
)

holdout_rate_summary <- data.frame(
  quantity = c("Observed", "Hurdle GAM", "Neuron baseline"),
  mean_event_rate = c(
    mean(test_predictions$event),
    mean(test_predictions$event_probability),
    mean(test_predictions$baseline_event_probability)
  )
)
knitr::kable(
  holdout_rate_summary,
  digits = 5,
  caption = "Mean event rate in the terminal holdout"
)
```

For event occurrence, higher AUC is better, while lower Brier score and log loss are better. AUC measures ranking but not probability calibration; Brier score and log loss are proper scoring rules. For positive and thresholded amplitudes, lower RMSE and MAE are better. The thresholded-amplitude prediction $\widehat p_{it}\widehat\mu_{it}$ evaluates both hurdle components jointly.

The out-of-sample comparison reverses the small training advantage. Event AUC
falls to
`r sprintf("%.3f", get_metric(test_metrics, "Event occurrence", "Hurdle GAM", "AUC"))`
and is below the neuron-only value of
`r sprintf("%.3f", get_metric(test_metrics, "Event occurrence", "Neuron baseline", "AUC"))`.
The GAM also has worse log loss
(`r sprintf("%.5f", get_metric(test_metrics, "Event occurrence", "Hurdle GAM", "log loss"))`
versus
`r sprintf("%.5f", get_metric(test_metrics, "Event occurrence", "Neuron baseline", "log loss"))`)
and worse Brier score. Conditional positive-amplitude RMSE is
`r sprintf("%.2f", get_metric(test_metrics, "Positive amplitude", "Hurdle GAM", "RMSE"))`
for the GAM versus
`r sprintf("%.2f", get_metric(test_metrics, "Positive amplitude", "Neuron baseline", "RMSE"))`
for the baseline. For the joint thresholded response, MAE is
`r sprintf("%.3f", get_metric(test_metrics, "Thresholded amplitude", "Hurdle GAM", "MAE"))`
versus
`r sprintf("%.3f", get_metric(test_metrics, "Thresholded amplitude", "Neuron baseline", "MAE"))`.
The behavioral model therefore loses to the simpler baseline on every reported
terminal-holdout metric.

```{r}
#| label: test-calibration
#| fig-cap: "Out-of-sample event-probability calibration in the terminal holdout."

test_calibration <- calibration_table(
  test_predictions$event,
  test_predictions$event_probability,
  n_bins = 10L
)

ggplot(test_calibration, aes(predicted, observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey45") +
  geom_point(aes(size = n), color = "#D55E00") +
  geom_line(color = "#D55E00") +
  coord_equal() +
  labs(x = "Mean predicted probability", y = "Observed event frequency", size = "Rows")
```

The holdout calibration curve lies mostly below the identity line, especially
for the highest predicted probabilities. On average, the GAM predicts an event
rate of
`r sprintf("%.4f", mean(test_predictions$event_probability))`, compared with
the observed rate of `r sprintf("%.4f", mean(test_predictions$event))`.
This systematic overprediction is consistent with the lower event prevalence
in the holdout (`r sprintf("%.4f", mean(test_data$event))`) than in training
(`r sprintf("%.4f", mean(train_data$event))`).

```{r}
#| label: test-time-prediction
#| fig-cap: "Observed and out-of-sample predicted population event rates in the terminal holdout."

test_time <- population_time_summary(test_predictions, bin_seconds = 1)
test_time_long <- rbind(
  data.frame(time_sec = test_time$time_sec, rate = test_time$observed_event_rate,
             series = "Observed"),
  data.frame(time_sec = test_time$time_sec, rate = test_time$predicted_event_rate,
             series = "Predicted")
)

ggplot(test_time_long, aes(time_sec, rate, color = series)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c(Observed = "black", Predicted = "#D55E00")) +
  labs(x = "Trial time (seconds)", y = "Mean event rate", color = NULL)
```

The time plot shows that overprediction persists across nearly the entire
holdout and becomes particularly large around 720--723 seconds. The smooth of
absolute time is evaluated just beyond its training range here; GAM smooths are
not reliable extrapolators, and its terminal trend appears to contribute to the
failure.

# Overall assessment

Generally speaking, this version of the model is **not yet a good predictive
model**. It is well calibrated in sample and the positive-amplitude component
has substantial in-sample deviance explained, but these facts do not survive
the more important terminal-block comparison. The neuron-only baseline is
better on all holdout metrics. The event AUC of approximately
`r sprintf("%.2f", get_metric(test_metrics, "Event occurrence", "Hurdle GAM", "AUC"))`
is only modestly above random ranking, and the GAM systematically overpredicts
the lower activity level at the end of the trial.

The results nevertheless identify several useful signals and failure modes:

1. Stable differences among neurons matter. The neuron-only baseline is strong,
   and neuron random intercepts account for an important part of the apparent
   amplitude fit.
2. A shared behavioral response adds a small amount in sample but is not stable
   at the end of the recording.
3. One population-wide position surface is biologically too restrictive for
   CA1 place cells, which can have distinct place fields.
4. The separate position and boundary effects are not identifiable in the
   current specification because boundary distance is almost completely
   concurvitable with two-dimensional position.
5. The smooth temporal trend misses rapid population fluctuations and
   extrapolates poorly into a terminal holdout.

These conclusions concern prediction within this trial. They do not imply that
mouse position or anatomical organization is biologically unimportant; rather,
the current pooled parameterization does not represent their neuron-specific
effects adequately.

# How to improve the model

## Simplify and validate the population-level terms

The first step should be an ablation analysis using several rolling-origin time
splits, not just the final 30 seconds. Compare the neuron baseline against
nested models containing speed, position, boundary distance, and time. In
particular:

- compare `s(x_z, y_z)` against `s(border_z)` instead of including both;
- remove `s(time_z)` initially, or represent time using coarse training blocks
  that do not require smooth extrapolation;
- report precision--recall performance in addition to ROC AUC because events
  occur in only about 0.5% of neuron-frame rows;
- evaluate several event thresholds and a lognormal or Gamma model for the
  excess amplitude $S_{it}-c_i$, since the observed positive distribution is
  threshold-truncated;
- use a block bootstrap or block-level validation for uncertainty because
  adjacent frames are dependent.

The preferred specification should be chosen from mean log loss across several
held-out time blocks. The final block can then remain untouched for a last
model comparison.

## Allow behavioral tuning to differ among neurons

The current random intercept allows different baseline rates but forces every
neuron to share the same position, boundary, speed, and time effects. This is
probably the central modeling limitation. A full two-dimensional smooth for
each of 462 neurons is too large for a first laptop model, but useful
intermediate choices include:

1. cluster neurons using training data only and fit group-specific position
   smooths, for example `s(x_z, y_z, by = neuron_group)`;
2. fit neuron-specific spatial models only for neurons with enough events and
   use a simpler partial-pooling model for sparse neurons;
3. use a reduced-rank hierarchical model in which neuron-specific spatial
   surfaces are combinations of a small number of shared spatial basis
   functions.

Any response-based clustering must be performed independently within each
training fold. Clustering once on the complete trial would leak holdout activity
into model construction.

## Use hippocampal neuron centroids

Let ${\bf c}_i=(c_{xi},c_{yi})$ denote neuron $i$'s centroid in the miniscope
field of view. The centroids describe anatomical neuron location, whereas
$(x_t,y_t)$ describes the mouse's position in the arena; these are distinct
coordinate systems and should never be combined as if they were one spatial
map.

The simplest anatomical model decomposes the neuron baseline into a smooth
centroid effect and residual neuron heterogeneity:

$$
b_i=h(c_{xi},c_{yi})+\epsilon_i.
$$

An exploratory `mgcv` implementation is

```r
event ~
  s(centroid_x_z, centroid_y_z, bs = "gp", k = 50) +
  s(neuron, bs = "re") +
  s(x_z, y_z, k = 40) +
  s(speed_z, k = 8)
```

The Gaussian-process centroid smooth asks whether nearby neurons have similar
baseline event propensities, while the neuron random effect retains residual
cell-specific variation. Because centroid coordinates are constant within a
neuron, these two terms can themselves be concurvitable; their penalties and
blocked predictive comparison must be examined. A more explicit hierarchical
version would give the vector of neuron effects a covariance that decays with
centroid distance:

$$
\operatorname{Cov}(b_i,b_j)=
\sigma_c^2\exp\{-\lVert {\bf c}_i-{\bf c}_j\rVert/\rho\}
+\sigma_i^2\mathbb 1(i=j).
$$

For prediction of these same 462 neurons, a centroid main effect may not beat
individual random intercepts: the random intercept already learns one baseline
per cell. Centroids become more valuable for structured shrinkage, prediction
for sparsely active or new neurons, and modeling how neuronal activity is
coordinated.

A practical next model is therefore to form anatomical groups using centroid
coordinates alone and allow group-specific mouse-position smooths:

$$
\operatorname{logit}(p_{it})=
\beta_0+b_i+f_{G_i}(x_t,y_t)+f_v(v_t),
$$

where $G_i$ is an anatomical group. Since the groups use centroids rather than
the held-out responses, this avoids response leakage. It directly tests whether
anatomically nearby neurons share behavioral tuning.

Finally, centroid distances can support a constrained Hawkes-inspired feature:

$$
A_{it}^{(\ell)}=
\sum_{j\ne i}\exp(-d_{ij}/\rho)Y_{j,t-\ell},
$$

with a few prespecified lags $\ell$. Adding these distance-weighted recent
population-activity summaries to the occurrence GAM would test whether nearby
neuronal activity improves prediction beyond mouse position, speed, and neuron
baseline. It is much smaller than estimating all $462^2$ directed interactions
and should still be interpreted as predictive coactivation, not synaptic
causality.

## Recommended sequence

The next analysis should proceed in this order:

1. establish a rolling-block baseline and remove the unstable absolute-time
   smooth;
2. choose either mouse position or boundary distance by blocked log loss;
3. add centroid-structured neuron baselines;
4. add centroid-defined group-specific place fields;
5. only then test distance-weighted lagged population activity.

Each step should be retained only if it improves prediction across multiple
held-out blocks rather than merely increasing in-sample deviance explained.
